%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,svgnames]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

% sorted
\usepackage{xcolor}
\usepackage[linesnumbered,vlined]{algorithm2e}
\usepackage{algorithmic}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel} 
\usepackage{csvsimple}
\usepackage{color}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{esvect}
\usepackage{hyperref}
%\usepackage{makeidx}
%\usepackage{mathpartir}
\usepackage{multicol}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage[ligature, inference]{semantic}
%\usepackage{subfig}
\usepackage{tikz}
\usepackage{xargs}
%\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{chngpage}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{colortbl}
\usepackage[most]{tcolorbox}

\input{defs.tex}


%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[FM@Lorentz]{A Research Agenda for Formal Methods in the
  Netherlands}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption




%% Title information
\title{A Research Agenda for Formal Methods in the Netherlands}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.


%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Marieke Huisman}
\affiliation{
  \institution{University of Twente}            %% \institution is required
}
\email{m.huisman@utwente.nl}          %% \email is recommended
%\and
\author{Eelco Visser}
\affiliation{
\institution{Technical University of Delft}
}
\email{e.visser@tud.nl}

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
On September 3 and 4, 2018, we organized a meeting on formal methods
research in the Netherlands. Goal of the meeting was to create a Dutch
formal methods community, to increase awareness of each other's
activities, and to find common grounds for collaborations. All
researchers working on formal methods in the Netherlands were invited
to contribute a 2-page abstract with their vision on the future of
formal methods research. This document bundles these visions.
\end{abstract}


% %% 2012 ACM Computing Classification System (CSS) concepts
% %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10011007.10011006.10011008</concept_id>
% <concept_desc>Software and its engineering~General programming languages</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003521.10003525</concept_id>
% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
% %% End of generated code


%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
%\maketitle


%% Bibliography
%\bibliography{bibfile}




%\bibliographystyle{plainurl}
%\pagestyle{plain}

\begin{document}

\maketitle

\section{Goal of the Workshop}
Research on Formal methods in the Netherlands is doing well. Success stories on the use of mechanized theorem proving, proof assistants, model checking, high-level programming languages, program verification etc. are ample. Our techniques are becoming mature and have the potential to be applied in industry.

However, research on formal methods is not ``done'' yet: there are still many interesting challenges ahead, for example concerning scalability and usability of formal methods. To be able to continue along this path, we need to improve visibility and funding opportunities for formal methods research.

To achieve this, we need a strong formal methods research community, that actively collaborates to lobby for research in this area.


To build up this community, we organized a two day workshop in the
Lorentz Center in Leiden on Monday 3, and Tuesday 4 September
2018. During this workshop, we brought together the key scientists in formal methods in The Netherlands.

The purpose of this meeting was to better understand our individual
goals and challenges, and to discuss about the potential and ideas for
research in formal methods for the next 10 years.  Goal of the
workshop was to achieve a better understanding of each other's
research areas, and to find common challenges and grounds for
collaborations, as such collaborations can further strengthen the
Dutch formal methods research community. Furthermore, we also believe
that being better informed about each other's research is a first step
towards creating a community with active lobbying power.

\section{Workshop Contents}
The workshop was filled with short presentations about current and
future research, with ample time for discussions. During the workshop,
presentations were given by: Marieke Huisman (UT), Jan Friso Groote
(TU/e), Jurriaan  Hage (UU), Marko  van Eekelen (OU), Eelco  Visser
(TUD), Robbert  Krebbers (TUD), Sung-Shik  Jongmans (OU), Jorge
 A. Perez (RUG), Frank  S. de Boer (UL/CWI), Wouter  Swierstra (UU), Sebastiaan
 Joosten (UT), Erik  Poll (RU), Gabriele Keller (UU), Sebastian
 Erdweg (TUD), Herbert  Bos (VU), Anton  Wijs (TU/e), 
Anna-Lena
 Lamprecht (UU),
 Jeroen Keiren (OU), Michel Reniers (TU/e), and Herman Geuvers (RU).



\section{This Document}
All participants
were asked to contribute a two page abstract with their vision for the
future of formal methods research. This document bundles the visions
of all participants, as well as the vision of some researchers in
formal methods in Netherlands that were not able to join the meeting. 







\end{document}
